---
title: "Stat 110: R Section "
author: Justin Zhu
subtitle: "Credit to Joe Blitzstein,  Kenneth Baclawski, & Matt DosSantos DiSorbo"
fontsize: 12pt
geometry: margin=1in
output: pdf_document
urlcolor: blue
---

# Good packages
```{r}
#install.packages('Rlab') 
library(Rlab)
```

# Bernoulli

```{r}
# Parameters (n, p for binomial and geometric)
set.seed(110)
n = 10
p.bern = 5/10
Y = rbern(n, p.bern)
Y

# Finding mean and variance
mean(Y)
var(Y)

# Graphics
# Overall histogram
hist(Y, main = "Y ~ Bern(n, p)", ylab = "Density", xlab = "Values", 
     col = rgb(0, 0, 0.5))
```

# Binomial

```{r}
# Parameters (n, p for binomial and geometric)
set.seed(110)
sims = 1000
n = 10
p.binom = 5/10
Y_binom = rbinom(sims, n, p.binom)

# Finding mean and variance
mean(Y_binom)
var(Y_binom)

# Graphics
# Overall histogram
#hist(Y_binom, main = "Y ~ Bern(n, p)", ylab = "Density", xlab = "Values", 
     #col = rgb(0.5, 0, 1, 1))
```
# What is the probability of getting at most 2?

```{r}
#Sum up the probability mass function!
dbinom(0, n, p.binom) + dbinom(1, n, p.binom) + dbinom(2, n, p.binom)

# Or alternatively, use the cumulative function:
pbinom(2, n, p.binom)
```

#Bridging the distributions

```{r}
set.seed(110)
sims = 1000
n = 10
p.binom = 5/10
Y_sum = rep(0, sims)
for (i in 1:sims){
  Y = rbern(n, p.bern)
  Y_sum[i] = sum(Y)
}

#hist(Y_sum, main = "Y_sum ~ n * Bern(p)", ylab = "Density", xlab = "Values", 
     #col = rgb(0.5, 0, 1, 1))

# We can compare the mean and variance between these two distributions
mean(Y_sum)
mean(Y_binom)
var(Y_sum)
var(Y_binom)
```

# Geometric
```{r}
set.seed(110)
# Same old, Same old: Parameters first!
n = 10
p.geom = 5/10

# Find the distribution
Y_geom = rgeom(n, p.geom)

# We can now plot
hist(Y_geom, main = "Y ~ Geom(n, p)", ylab = "Density", xlab = "Values", 
     col = rgb(0.5, 0, 1, 1))
```

# Finding probabilities
```{r}
seven_heads = dgeom(7, p.geom)
seven_heads
```

# Hypergeometric

Think of a jar with  b  blue balls and  w  white balls. If you are drawing  n  balls total and are hoping to pick the white balls and not pick the blue balls and you then  $X$  be the number of white balls that you pick, then  $X$  has a Hypergeometric distribution.  HGeom(w,b,n).

```{r}
# white balls
m = w = 8
n = b = 5
k = 3
x = 2

dhyper(x, m, n, k)
```

