---
title: "Stat 110: Section 1"
author: Justin Zhu
date: August 25, 2018
subtitle: "Adapted from Joe Blitzstein and Jessica Hwang's Introduction to Probability and Kenneth Baclawski's Introduction to Probability with R"
fontsize: 12pt
geometry: margin=1in
output: pdf_document
urlcolor: blue
---

# Getting familiar with R

You can follow along by typing the scripts into a console or an R file in RStudio IDE.  This R Markdown notebook will contain the R script on one line and the subsequent output in the next line.  You can press "Run" to run the selected lines while interacting with this RMarkdown file.

For example,
```{r}
1+1
```

The R code is "1+1" and the output looks something like "##[1] 2".

Some more scripts to play around with:
```{r}
2 * pi * 5 + 100
100 * 299792458 ^ 2
```

R follows conventional use of mathematical operators and order of operations.

# R is built around vectors.  
A simple vector is the sequence 1, 2, . . ., n. If you want n to be 100, then we type this command:
```{r}
1:100
```

We also use the ```c``` command to combine and concactenate values into a vector.

```{r}
c(1,4,9,16,25,36)
```

Here, we stored those values into v.  
```{r}
v <- c(1,4,9,16,25,36)
```
We can access elements inside v using the following:

```{r}
v[2]
```

We can also find several values within v an form a new vector
```{r}
v[c(1,3,5)]
```

R does *not* use zero-index.

# Variables and Functions
We have already seen how "<-" is used in R to store values into a variable.
You can also use "->" in the other direction.  The arrow, however must point to the variable is assigned the value.  For example, in this example, triplepie and pietriple are both being assigned the same values.

```{r}
triplepie <- 3*pi

3*pi -> pietriple
```

We can use boolean operators to verify whether certain variables contain the same values
```{r}
triplepie == pietriple
```

The boolean operators in R also follow conventional programming language practice.  You can find a list of them [here](https://www.statmethods.net/management/operators.html)

To create a function we can use the "function" keyword and specify parameters.  We can write a function that calculates the probability of at least one birthday match in a group of n people.

```{r}
birthday <- function(n) {
  days_in_year <- 365
  ans <- 1-prod((days_in_year-n+1):days_in_year)/days_in_year^n
  return(ans)
}
```

Here, ```prod``` returns the product of all the values passed in as arguments.

You can always use help() to find more information about R's built-in functions.

```{r}
help(prod)
```

We can now finally celebrate some birthdays by typing our function name with the specified parameters!

```{r}
birthday(1)
birthday(23)
```

How can we plot the relationship between number of people and the probability of getting a matching birthday?  How does changing the number of people in our birthday problem affect the probability?


```{r}
num_of_people <- 1:50
matching_birthdays_probability <- lapply(num_of_people, birthday)
plot(num_of_people,matching_birthdays_probability)
```







